Title Goes Here \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2021-04-06 analysis/paper/paper.tex
Muchos cambios

\usepackage{gensymb}
\usepackage{subfig}
\usepackage[inline]{showlabels}
\usepackage{chngcntr}
\usepackage{natbib}
\usepackage{lineno}
\linenumbers
\usepackage{pdfcomment}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\institute{
        Elio Campitelli \at
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
     \email{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Leandro B. Díaz \at
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Carolina Vera \at
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
CONICET -- Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
CNRS -- IRD -- CONICET -- UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

\pdfcomment[]{¿Agrego algo sobre impactos? Por ejemplo, impactos de PSA /SAM.}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone COlumn (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{era5} for the period 1979 to 2019. We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends. We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and computed horizontal wave activity fluxes following \citet{plumb1985} (Equation 5.7). For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}.

We computed the amplitude and phase of the zonal wave 1 by averaging variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) 70°S and 45°S for each Spring and computing the Fourier spectrum.

We computed the level-dependent Southern Annular Mode index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}. We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively). These are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields \citep{campitelli2021}. Since the analysis here is only for the SON trimester, monthly values were averaged across trimesters.

@@

\hypertarget{complex-empirical-orthogonal-functions}{%
\subsection{Complex Empirical Orthogonal Functions}\label{complex-empirical-orthogonal-functions}}

\begin{figure}
\centering
\includegraphics{../figures/eof-naive-1.pdf}
\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units). The numbers at the top-left of each panel ate the zonal variance explained by each EOF.}
\end{figure}

In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase. For instance, Figure~\ref{fig:eof-naive} shows the leading 4 EOFs of zonally anomalous September through October geopotential height at 50 hPa and 100 hPa. It is clear that at 50 hPa the first two EFS represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3 pattern shifted by 1/4 wavelength. Since these pairs of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise these waves \citep{horel1984}. This methods involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it. In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time. Since the latitude circle is a periodic domain, this procedure does not suffer from edge effects.

The result of the cEOF is a set of complex spatial patterns and complex time series. The real and imaginary part of each spatial pattern represent two phases wave-like spatial pattern that are in quadrature. The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.

@@

\caption{\label{tab:corr-ceof-splitted}$R^2$ of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.}
\centering
\begin{tabular}[t]{l>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{} & \multicolumn{3}{c}{50 hPa} \\
\cmidrule(l{3pt}r{3pt}){2-4}
200 hPa & PC1 & PC2 & PC3\\
\midrule
PC1 & \cellcolor[HTML]{E1C0BB}{\textcolor{black}{0.28}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
PC2 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{BB7A73}{\textcolor{white}{0.60}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
PC3 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FDFCFC}{\textcolor{black}{0.01}}\\
\bottomrule
\end{tabular}
\end{table}

Another characteristic seen in Figure~\ref{fig:eof-naive} is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa. The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa. Furthermore, Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF across levels. There's a high degree of correlation between the respective EOF1 and EOF2 at each level. Both the spatial similarities and temporal correlations suggest some level of joint variability between levels.

Both observations motivate the decision of performing complex EOF jointly between levels. Since the variability of the stratosphere is far greater in magnitude than the variability of the troposphere, we standardised the values by the standard deviation of each level. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time.

The phase of principal components is defined up to an additive constant. For real principal components, this constant can be either 0 or \(\pi\), corresponding to a change in sign. For complex principal components, it can be any number between 0 and \(2\pi\) \citep{horel1984}. Since any choice is arbitrary and equally valid, we chose the phase of each EOF so that the real and imaginary parts are aligned with meaningful phases in our analysis. This was done after some exploratory analysis.

\begin{figure}
\centering
\includegraphics{../figures/rotations-1.pdf}
\caption{\label{fig:rotations}a) Correlation of the Real and Imaginary EOF1 with wave 1 amplitude of Total Ozone Column averaged between 70°S and 45°S and b) \(R^2\) between the Real and Imaginary EOF2 and the ONI index for different values of rotation. The vertical line marks the selected rotation, which maximises de correlation.}
\end{figure}

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 70°S and 45°S. This also nearly minimises the correlation with the imaginary part (Figure~\ref{fig:rotations}a).

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure~\ref{fig:rotations}b). It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase. If the relationship is very weak, the rotation cannot generate a strong relationship.

@@

\hypertarget{description-of-eofs}{%
\subsection{Description of EOFs}\label{description-of-eofs}}

\begin{figure}
\centering
\includegraphics{../figures/ceofs-1-1.pdf}
\caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. Real part in shading, imaginary part in contours}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/ceofs-2-1.pdf}
\caption{\label{fig:ceofs-2}Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.}
\end{figure}

Figures \ref{fig:ceofs-1} and \ref{fig:ceofs-2} show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa. In the spatial patterns in Figure~\ref{fig:ceofs-1}, the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the complex temporal EOF.

The wave patterns described by these complex EOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}: The first is a wave 1, while the second is a wave 3. Note that in Figure~\ref{fig:ceofs-2}, both parts of the EOF1 have non zero mean. This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean, not the time mean. The variability associated with the first EOF includes variability that projects onto the mean zonally anomalous field.

@@

Figure~\ref{fig:extended-series} shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.There is an upward trend in the real part of EOF1 (Figure~\ref{fig:extended-series}a.1, p-value~ \textless{} 0.001) and no significant trend in any of the complex parts of EOF2. The positive trend in the Real EOF1 translates into a positive trend in EOF1 magnitude, but not in phase (not shown).

@@

The spatial patterns shown in Figure~\ref{fig:ceofs-1} are ``pure'' EOF patterns. But as they are derived by removing the zonally symmetric circulation, they might not include all the variability that is actually associated with the EOF time series. They are also idealised in that they are forced by construction, to be Hilber transforms of each other. To get a more realistic view of the real-world variability described by these EOFS, we computed regression patterns of each EOF with the complete geopotential fields.

@@

Figure~\ref{fig:eof1-regr-gh} shows regression patterns of EOF1 and geopotential height. At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase. Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the Real EOF1 located towards the dateline, and the one of the Imaginary EOF1 located over Eastern Antarctica. However, the Real EOF1 pattern is substantially altered by the zonally symmetric circulation. Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

Similarly affected are the regression patterns at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b). It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure~\ref{fig:eof1-regr-gh}b.1). The Imaginary EOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.

With the exception of the Imaginary EFO1, it's clear that these patterns are very different than the idealised versions shown in Figure~\ref{fig:ceofs-1}, particularly at 200 hPa. Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º. This should not be utterly surprising, since the spatial patterns of the \emph{naive} EOFs only show this feature at 50 hPa, suggesting that using the complex EOF method is ``forcing'' a wave structure where there is none.

Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere. While in the troposphere they are associated with slightly off-centre monopoles.

@@

Figure~\ref{fig:eof2-regr-gh} shows the regression pattern of geopotential height and the EOF2. Unlike in the case of EOF1, in this case the regression patterns represent relatively well the \emph{idealised} patterns from Figure~\ref{fig:ceofs-1}. Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real EOF2 are 90º out of phase with those associated with the Imaginary EOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere.

EOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns \citep{mo2001}. Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figure 1 a and b from \citet{mo2001}, the Real EOF2 can roughly be identified with the PSA2, while the Imaginary EOF2 resembles PSA1. This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability.

\hypertarget{relationship-with-other-variables}{%
\subsection{Relationship with other variables}\label{relationship-with-other-variables}}

\hypertarget{ozone}{%
\subsubsection{Ozone}\label{ozone}}

@@

\begin{figure}
\centering
\includegraphics{../figures/t-vertical-1.pdf}
\caption{\label{fig:t-vertical}Regression of EOF1 with zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 25ºS for the period 1979 -- 2019.}
\end{figure}

Figure~\ref{fig:eof1-regr-t} shows regression patterns of EOF1 with air temperature. In both levels, the distribution of temperature anomalies mirror the geopotential height anomalies in 50 hPa. To understand the vertical distribution of these anomalies, Figure~\ref{fig:t-vertical}) shows the regression pattern of EOF1 with zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS. Temperature zonal anomalies associated with EOF1 show a clear wave 1 pattern throughout the atmosphere above 200 hPa with a change in sign above 10 hPa. Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa; that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere. This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies \citep{hartmann1979, wirth1993, smith1995}.

\begin{figure}
\centering
\includegraphics{../figures/o3-regr-1.pdf}
\caption{\label{fig:o3-regr}Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column. On contours, the mean zonal anomaly of Total Ozone Column (negative contours in dashed lines). For the 1979 -- 2019 period.}
\end{figure}

>>>>

\usepackage{gensymb}
\usepackage{subfig}
\usepackage[inline]{showlabels}
\usepackage{chngcntr}
\usepackage{natbib}
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\institute{
        Elio Campitelli \at
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
     \email{\href{mailto:elio.campitelli@cima.fcen.uba.ar}{\nolinkurl{elio.campitelli@cima.fcen.uba.ar}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Leandro B. Díaz \at
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Carolina Vera \at
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos. Buenos Aires, Argentina.
CONICET -- Universidad de Buenos Aires. Centro de Investigaciones del Mar y la Atmósfera (CIMA). Buenos Aires, Argentina.
CNRS -- IRD -- CONICET -- UBA. Instituto Franco-Argentino para el Estudio del Clima y sus Impactos (IRL 3351 IFAECI). Buenos Aires, Argentina. \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

\hypertarget{data-and-methods}{%
\section{Data and Methods}\label{data-and-methods}}

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone Column (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 \citep{era5} for the period 1979 to 2019. We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 \citep{era5be} to look at long-term trends. We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK \citep{fishpack} and computed horizontal wave activity fluxes following \citet{plumb1985}. For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation \citep{cmap}, with a 2.5\degree~resolution in latitude and longitude. This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

We restrict our study to the September-October-November (SON) trimester. For that, we use seasonal means, averaging monthly values weighted by the number of days in each month. We also use 200 hPa to represent the high troposphere and 50 hPa to represent the lower stratosphere.

We computed the amplitude and phase of the zonal wave 1 by averaging (area-weighted) variables (temperature, geopotential height, ozone mixing ratio and Total Ozone Column) between 70°S and 45°S for each SON and extracting the wave-1 component of the Fourier spectrum. We chose this latitude band because it is wide enough to capture most of the relevant anomalies.

We computed the level-dependent Southern Annular Mode (SAM) index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period \citep{baldwin2009}. We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively). These are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields \citep{campitelli2021}. Since the analysis here is only for the SON trimester, monthly values were averaged across trimesters.

@@

\textcolor{red}{Creo que habría que expandir un poco esta parte e ir comentando todoas las metodologías que vas usando: regresiones, tendencias, cómo evlauas significancia en cada caso, etc. }

\hypertarget{complex-empirical-orthogonal-functions-ceof}{%
\subsection{Complex Empirical Orthogonal Functions (cEOF)}\label{complex-empirical-orthogonal-functions-ceof}}

\begin{figure}
\centering
\includegraphics{../figures/eof-naive-1.pdf}
\caption{\label{fig:eof-naive}Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units). The numbers at the top-left of each panel ate the zonal variance explained by each EOF.}
\end{figure}

In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase \citep{horel1984}. For instance, Figure~\ref{fig:eof-naive} shows the leading 4 EOFs of zonally anomalous SON geopotential height at 50 hPa and 200 hPa. It is clear that at 50 hPa the first two EOFs represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave pattern with shorter wavenumber shifted by 1/4 wavelength. A similar EOF structure can be seen in 200 hPa (not shown). Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise these waves \citep{horel1984}. This method involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it. In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time. Since the latitude circle is a periodic domain, this procedure does not suffer from edge effects. The result of the cEOF is a set of complex spatial patterns and complex time series. The real and imaginary part of each spatial pattern represent two phases wave-like spatial pattern that are in quadrature. The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.

@@

\caption{\label{tab:corr-ceof-splitted}Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.}
\centering
\begin{tabular}[t]{l>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{} & \multicolumn{3}{c}{50 hPa} \\
\cmidrule(l{3pt}r{3pt}){2-4}
200 hPa & cEOF1 & cEOF2 & cEOF3\\
\midrule
cEOF1 & \cellcolor[HTML]{E1C0BB}{\textcolor{black}{0.28}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
cEOF2 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{BB7A73}{\textcolor{white}{0.60}} & \cellcolor[HTML]{FDFAFA}{\textcolor{black}{0.02}}\\
cEOF3 & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FFFFFF}{\textcolor{black}{0.00}} & \cellcolor[HTML]{FDFCFC}{\textcolor{black}{0.01}}\\
\bottomrule
\end{tabular}
\end{table}

Table \ref{tab:corr-ceof-splitted} shows the coefficient of determination between time series of the amplitude of each complex EOF (cEOF) across levels. There's a high degree of correlation between the respective cEOF1 and cEOF2 at each level. The spatial patterns of the 50 hPa and 200 hPa cEOFs are also similar (not shown)

Both the spatial pattern similarity and the high temporal correlation of cOEFs computed at 50 hPa and 200 hPa suggest that these are, to a large extent, modes of joint variability. This motivates the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

The phase of principal components is defined up to an additive constant. For real principal components, this constant can be either 0 or \(\pi\), corresponding to a change in sign. For complex principal components, it can be any number between 0 and \(2\pi\) \citep{horel1984}. Since any choice is arbitrary and equally valid, we chose the phase of each EOF so that the real and imaginary parts are aligned with meaningful phases in our analysis. This procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.

For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 70°S and 45°S. This also nearly minimises the correlation with the imaginary part.

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index \citep{bamston1997} and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.

@@

\hypertarget{ceof-patterns}{%
\subsection{cEOF patterns}\label{ceof-patterns}}

\begin{figure}
\centering
\includegraphics{../figures/ceofs-1-1.pdf}
\caption{\label{fig:ceofs-1}Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019. Real part in shading, imaginary part in contours.}
\end{figure}

\begin{figure}
\centering
\includegraphics{../figures/ceofs-2-1.pdf}
\caption{\label{fig:ceofs-2}Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.}
\end{figure}

To describe the spatial and temporal variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa are shown in Figures \ref{fig:ceofs-1} and \ref{fig:ceofs-2}. The first mode (CEOF1) explains 82\% of the variance, while the second mode (CEOF1) explains a smaller fraction (7\%). In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the complex temporal EOF. The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}.

The first cEOF (Fig.~\ref{fig:eof-naive} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes. At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is locates at around 175ºE indicating a westerly shift in phase.

The second cEOF (Fig.~\ref{fig:eof-naive} column~b) show also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales. In particular, it's dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector. This modulated amplitude is especially apparent at 200 hPa. There is no apparent phase shift, suggesting a rather barotropic structure.

The temporal variability for both components of the cEOF1 have non zero mean (Fig.~\ref{fig:ceofs-2}). This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean. As a consequence, the variability associated with the first cEOF includes variability that projects onto the mean zonally anomalous field. There are no significant simultanous correlation between the time series.

@@

Both cEOF shot year-to-year variability but show no evidence of decadal variabilty. The relatively short timespan (40 years) hampers our habilty to detect longer trends. Therefore, we extend the cEOF time series using the preliminary ERA5 back extension going back to 1950 by projecting the older geopogential height zonal anomaly fields onto the spatial patterns shown in Figure~\ref{fig:ceofs-1}.

The extended cEOF timeseires are shown in Figure~\ref{fig:extended-series}. There is a significant upward trend in the real component of cEOF1 (Fig.~\ref{fig:extended-series}a.1, p-value~\textless{} 0.001) and no significant trend in any of the complex components of cEOF2. The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).

This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1. A similar observation was made by \citet{raphael2003}, who detected a step after around 1975 in the leading EOF of August-September-October 500 hPa zonal geopotential height anomalies.

@@

The spatial patterns shown in Figure~\ref{fig:ceofs-1} are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series. To understand the full geopotential height field changes associated with each cEOF, we computed regression patterns of each cEOF with the complete geopotential fields.

@@

Figure~\ref{fig:eof1-regr-gh} shows regression patterns between geopotential height anomalies and cEOF1. At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase. Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica. However, the cReal EOF1 pattern is substantially altered by the zonally symmetric circulation. Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

The regression patterns at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) are similarly influenced by the zonally symmetric circulation. It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure~\ref{fig:eof1-regr-gh}b.1). The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.

With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.~\ref{fig:ceofs-1}), particularly at 200 hPa. Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º. This should not be utterly surprising, since the spatial patterns of the \emph{naive} EOFs only show this feature at 50 hPa (Fig.~\ref{fig:eof-naive}), suggesting that using the cEOF method is artificially generating a wave structure at 200hPa. Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere. While in the troposphere, they are associated with slightly off-centre monopoles.

@@

Figure~\ref{fig:eof2-regr-gh} shows the regression pattern of geopotential height and the cEOF2. Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure~\ref{fig:ceofs-1}. Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans. cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns \citep{mo2001}. Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from \citet{mo2001}, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1.

\hypertarget{ceofs-relationship-with-other-variables}{%
\subsection{cEOFs relationship with other variables}\label{ceofs-relationship-with-other-variables}}

\hypertarget{temperature-and-ozone}{%
\subsubsection{Temperature and Ozone}\label{temperature-and-ozone}}

@@

\begin{figure}
\centering
\includegraphics{../figures/t-vertical-1.pdf}
\caption{\label{fig:t-vertical}Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 35ºS for the period 1979 -- 2019.}
\end{figure}

Figure~\ref{fig:eof1-regr-t} shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1. In both levels, the Real cEOF1 is associated with positive monopole over the South Pole with its centre moved slightly towards 150ºE (Fig.~\ref{fig:eof1-regr-t} column 1). Regression maps on the Imaginary cEOF1, on the other hand, show a more clear wae 1 pattern with its maximum around 60ºE. The distribution of temperature regression coefficients at 50 hPa and at 200 hPa mirror the geopotential height regression maps at 50 hPa (Fig.~\ref{fig:eof1-regr-gh}).

The vertical distribution of these patterns is shown in Figure~\ref{fig:t-vertical}) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS. Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 250 hPa with a change in sign above 10 hPa. Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.~\ref{fig:t-vertical})); that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere . This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere. In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies \citep{hartmann1979, wirth1993, smith1995}.
